# Gene sets

```{r echo = FALSE}
knitr::opts_chunk$set(cache = TRUE)
```

## What do we want to do?

Return to our top table

```{r}
toptbl <- readRDS("toptbl.Rda")
toptbl
```

One small correction to the top table is that the ENTREZID should be a
character() vector

```{r}
toptbl <- 
  toptbl %>% 
  mutate(ENTREZID = as.character(ENTREZID))
```

Biologically, genes always function in some kind of relation with other genes
and the biology of organsims. For instance, genes
are expressed in particularly cellular compartments, or participate in
particular biological processes. We would like to use established _ontologies_
that describe these relationships.

## Mapping from (entrez) gene identiifiers to GO identifiers

The [org.Hs.eg.db][] packages provides a way to map from (Entrez) gene identifiers
to GO terms.

```{r, message = FALSE}
library(org.Hs.eg.db)
```

Note that the keytypes of `org.Hs.eg.db` contain (Entrez) gene IDs, and the
columns contain GO ids

```{r}
keytypes(org.Hs.eg.db)
columns(org.Hs.eg.db)
```

Genes can belong to more than on GO category (e.g., participate in more than
one biological pathway). Use `AnnotationDbi::select()` to map between ENTREZID
and GO id; we'll represent the result as a tibble for convenience

```{r}
eggo <- AnnotationDbi::select(org.Hs.eg.db, toptbl$ENTREZID, "GO") %>%
  as_tibble()
eggo
```

Let's focus on the 'biological process' (BP) ontology, and not worry about
'EVIDENCE' (the type of evidence, e.g., experimental verification,
computational inference, ...) that supports the classification of a particular
ENTREZ identifier to a particular GO identifier.

```{r}
eggo <-
  eggo %>%
  dplyr::rename(GOID = "GO") %>%
  dplyr::select(-EVIDENCE) %>%
  dplyr::filter(ONTOLOGY %in% "BP") %>%
  dplyr::distinct()
eggo
```

[org.Hs.eg.db]: https://bioconductor.org/packages/org.Hs.eg.db

## Using GO (gene ontology) annotations to group genes by biological process

The [GO.db][] package provides information on GO identifiers.

```{r, message = FALSE}
library(GO.db)
columns(GO.db)
```

Use `AnnotationDbi::select()` to explore information about a particular GO id

```{r, message = FALSE}
goid <- "GO:0007165"
AnnotationDbi::select(GO.db, goid, "TERM", "GOID")
def = AnnotationDbi::select(GO.db, goid, "DEFINITION", "GOID")
cat(strwrap(def$DEFINITION), sep="\n")
```

Add terms and definitions to all rows of `eggo`

```{r}
goinfo <- 
  AnnotationDbi::select(
    GO.db, eggo$GOID, c("TERM", "DEFINITION"), "GOID"
  ) %>% 
  as_tibble() %>%
  distinct()
goinfo
```

Use `dplyr::left_join()` to join all rows of `eggo` to their corresponding
term and definition...

```{r}
eggoinfo <- left_join(eggo, goinfo, by = "GOID")
eggoinfo
```

...and even add the top table gene expression information

```{r}
left_join(eggoinfo, toptbl, by = "ENTREZID")
```

[GO.db]: https://bioconductor.org/packages/GO.db

## More flexible gene set annotation using [BiocSet][]

[BiocSet]: https://bioconductor.org/packages/BiocSet

## Saving and session information

```{r}
sessionInfo()
```
