# (Advanced) Working with SQL databases

```{r echo = FALSE}
knitr::opts_chunk$set(cache = TRUE)
```

## What do we want to do?

Return to our EnsDb package

```{r, message = FALSE, cache = FALSE}
library(EnsDb.Hsapiens.v75)
EnsDb.Hsapiens.v75
```

Notice the second line of output above

```
|Backend: SQLite
```

The package actually 'wraps' a SQLite data base. Databases are very useful for
managing large amounts of data, especially when the data consist of several
'tables' that are related to one another in various ways.

The [dbplyr][] package provides a very convenient way to work with databases,
especially if you know a little R and 'tidyverse', but only some SQL.

```{r, message = FALSE, cache = FALSE}
library(dbplyr)
```

[dbplyr]: https://cran.r-project.org/package=dbplyr

## Using [dbplyr][] to access the [EnsDb.Hsapiens.v75] database

_R_ connects to SQL databases use 'connections'. Various packages provide 
interfaces that connect to a number of different databases, e.g., [RSQLite][], [RMariaDB][], ... The databases can be local, or hosted on a server.

[RSQLite]: https://cran.r-project.org/package=RSQLite
[RMariaDB]: https://cran.r-project.org/package=RMariaDB

We obtain a connection to the database underlying our EnsDb object using
`dbconn()`; usually one would use the `dbConnect()` method defined in the corresponding

```{r}
dbconn(EnsDb.Hsapiens.v75)
```

Use `src_dbi()` to access the database connection from dbplyr

```{r, cache = FALSE}
tbls <- src_dbi( dbconn( EnsDb.Hsapiens.v75 ) )
tbls
```

The output shows the location of the sqlite database, as well as the tables
present in the database.

## Working with database tables

Many database operations are very straight-forward with dbplyr. Attach the
[dplyr][] package to our session

[dplyr]: https://cran.r-project.org/package=dply

```{r, message = FALSE}
library(dplyr)
```

Select the table of interest

```{r}
tbl(tbls, "gene")
```

Note that, unlike the output of an in-memory `tibble()`, the output does not
display the number of rows in the database table -- dbplyr has retrieved only
enough information from the database to display on the screen. 

Apply many standard dplyr verbs, e.g.,

```{r}
tbl(tbls, "gene") %>%
    dplyr::count(gene_biotype) %>%
    arrange(desc(n))

tbl(tbls, "gene") %>%
    dplyr::filter(gene_biotype %in% c("lincRNA"))
```

Here's the 'tx' table, which includes the relationship between transcript and
gene

```{r}
tbl(tbls, "tx") %>%
    dplyr::select(gene_id, tx_id)
```

And the 'protein' table, which relates transcript to protien

```{r}
tbl(tbls, "protein")
```

If we wished to know the protien sequences produced by different genes, we
could form a join between these two tables

```{r}
left_join(
    tbl(tbls, "protein"),
    tbl(tbls, "tx") %>%
        dplyr::select(gene_id, tx_id),
    by = "tx_id"
)
```

Knowledge of dplyr verbs coupled with basic knowledge of SQL databases provide
very useful tools for working with large, relationional data.

## Saving and session information

```{r}
sessionInfo()
```